---
title: "brfss_prop_tests"
author: "Shivalika Chavan"
date: "2025-11-23"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
knitr::opts_chunk$set(
  fig.width = 12, 
  fig.asp = 0.9,  
  out.width = "95%"
)
library(tidyverse)
library(patchwork)
library(here)

source(here("source", "load_clean_brfss.R"))
brfss_data = load_clean_brfss(here("data", "brfss_clean_2017_2024.csv.zip"))
state_fips = read_csv(here("data", "legal_sports_report", "state_fips.csv"))
state_fips_legalized_only = read_csv(here("data", "legal_sports_report", "state_legalization_dates.csv"))

outcome_vars <- c("any_physical_health_not_good_days", "any_mental_health_not_good_days", "has_depressive_disorder", "has_binge_drink")
```

## Prop Test and Plotting Functions

Writing a function for running prop test by year (2024 vs 2017)
```{r}
run_prop_test_state = function(state_df){
  
  counts = 
    state_df |> 
    group_by(year = year(date), outcome_value) |>
    summarize(
      n = n(), 
      .groups = "drop"
      ) |> 
    filter(!is.na(outcome_value)) |> 
    pivot_wider(
      names_from = outcome_value,
      values_from = n
      ) |> 
    mutate(total_responses = `TRUE` + `FALSE`) 
  
  x_poor_outcome <- counts |> pull(`TRUE`)
  n_total_counts <- counts |> pull(total_responses)
  
  if (nrow(counts) < 2) {
    return(tibble(NA))
  }
  
  # alternative is less since we are testing that 2017 < 2024
  prop_test_result = prop.test(
    x = x_poor_outcome, 
    n = n_total_counts, 
    alternative = "less", 
    conf.level = 0.95
  ) 
  
  prop_test_result |> 
    broom::tidy()
    
}
```

Writing function for running prop test using `sb_legal` as the inflection point, instead of 2024 vs 2017
```{r}
run_prop_test_state_legalization = function(state_df){
  
  counts = 
    state_df |> 
    group_by(sb_legal, outcome_value) |>
    summarize(
      n = n(), 
      .groups = "drop"
      ) |> 
    filter(!is.na(outcome_value)) |> 
    pivot_wider(
      names_from = outcome_value,
      values_from = n
      ) |> 
    mutate(total_responses = `TRUE` + `FALSE`) 
  
  x_poor_outcome <- counts |> pull(`TRUE`)
  n_total_counts <- counts |> pull(total_responses)
  
  if (nrow(counts) < 2) {
    return(tibble(NA))
  }
  
  prop_test_result = prop.test(
    x = x_poor_outcome, 
    n = n_total_counts, 
    alternative = "less", 
    conf.level = 0.95
  ) 
  
  prop_test_result |> 
    broom::tidy()
    
}
```

Writing a function for plotting (consolidated into 1)
```{r}
plot_outcome_props = function(outcome, df){
  
  title_map = c(
      "any_physical_health_not_good_days" = "At Least 1 Not Good Physical Health Day",
      "any_mental_health_not_good_days" = "At Least 1 Not Good Mental Health Day",
      "has_depressive_disorder" = "Has Depressive Disorder",
      "has_binge_drink" = "Has Binge Drink"
    )
  
  plot_title = title_map[outcome]
  
  p = df |> 
    ggplot(aes(y = fct_reorder(abbr, estimate1))) + 
    geom_point(aes(x = estimate1), color = "blue", size = 2) +
    geom_point(aes(x = estimate2), color = "red", size = 2) +
    geom_segment(aes(x = estimate1, xend = estimate2, color = significant_increase)) + 
    scale_color_manual(
      values = c("TRUE" = "red", "FALSE" = "gray"),
      labels = c("Significant Increase (p < 0.05)", "No Significant Increase"),
      name = "Change Magnitude"
    ) +
    xlim(0, 0.65) + 
    labs(
      title = plot_title,
      y = "State"
    ) +
    theme_minimal() +
    theme(legend.position = "none") 
    
  p
}
combine_plots = function(plotting_df){
  
  outcome_plots = 
    plotting_df |> 
    group_by(outcome) |> 
    nest() |> 
    mutate(
      plot = map2(outcome, data, plot_outcome_props)
    )
  
  plot_list = outcome_plots |> pull(plot)
  
  plots_combined = wrap_plots(plot_list, ncol = 2)
  
  plots_combined
    
}
```

## Comparisons

## 2017 vs. 2024
This includes all states in 2017 and 2024 and looking at the general trend in health outcomes irrespective of legalization status. 
```{r}
prop_tests_state = 
  brfss_data |> 
  filter(year(date) %in% c(2017, 2024)) |>
  select(state, date, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(test_result = map(data, run_prop_test_state)) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(significant_increase = p.value < 0.05)

prop_tests_state

plotting_df = 
  prop_tests_state |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))) 

combine_plots(plotting_df)
```

## Before and After Legalization
Now instead of just looking at the oldest and newest available data, let's use sb_legal as the inflection point. 
```{r}
prop_tests_sb_legal = 
  brfss_data |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(
    test_result = map(data, run_prop_test_state_legalization)
  ) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(
    significant_increase = p.value < 0.05
  )

prop_tests_sb_legal

plotting_df = 
  prop_tests_sb_legal |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(
    significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))
  ) 

combine_plots(plotting_df)
```

Going a step further, looking only at adults under 50, since a majority of sports bettors are younger. 
```{r}
prop_tests_sb_legal = 
  brfss_data |> 
  filter(age_group_5yr %in% c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49")) |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(
    test_result = map(data, run_prop_test_state_legalization)
  ) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(
    significant_increase = p.value < 0.05
  )

prop_tests_sb_legal

plotting_df = 
  prop_tests_sb_legal |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(
    significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))
  ) 

combine_plots(plotting_df)
```