---
title: "brfss_prop_tests"
author: "Shivalika Chavan"
date: "2025-11-23"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
knitr::opts_chunk$set(
  fig.width = 12, 
  fig.asp = 0.9,  
  out.width = "95%"
)
library(tidyverse)
library(patchwork)
library(here)

source(here("source", "load_clean_brfss.R"))
source(here("source", "prop_test_functions.R"))
brfss_data = load_clean_brfss(here("data", "brfss_clean_2017_2024.csv.zip"))
state_fips = read_csv(here("data", "legal_sports_report", "state_fips.csv"))
state_fips_legalized_only = read_csv(here("data", "legal_sports_report", "state_legalization_dates.csv"))

outcome_vars <- c("any_physical_health_not_good_days", "any_mental_health_not_good_days", "has_depressive_disorder", "has_binge_drink")
```

## Comparisons

## 2017 vs. 2024
This includes all states in 2017 and 2024 and looking at the general trend in health outcomes irrespective of legalization status. 
```{r}
prop_tests_state = 
  brfss_data |> 
  filter(year(date) %in% c(2017, 2024)) |>
  select(state, date, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(test_result = map(data, run_prop_test_state)) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(significant_increase = p.value < 0.05)

prop_tests_state

plotting_df = 
  prop_tests_state |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))) 

combine_plots(plotting_df)
```

## Before and After Legalization
Now instead of just looking at the oldest and newest available data, let's use sb_legal as the inflection point. 
```{r}
prop_tests_sb_legal = 
  brfss_data |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(
    test_result = map(data, run_prop_test_state_legalization)
  ) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(significant_increase = p.value < 0.05)

prop_tests_sb_legal

plotting_df = 
  prop_tests_sb_legal |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))) 

combine_plots(plotting_df)
```

Going a step further, looking only at adults under 50, since a majority of sports bettors are younger. 
```{r}
prop_tests_sb_legal = 
  brfss_data |> 
  filter(age_group_5yr %in% c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49")) |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(
    test_result = map(data, run_prop_test_state_legalization)
  ) |> 
  select(-data) |> 
  unnest(cols = c(test_result)) |> 
  select(state, outcome, estimate1, estimate2, p.value, conf.low, conf.high) |> 
  mutate(significant_increase = p.value < 0.05)

prop_tests_sb_legal

plotting_df = 
  prop_tests_sb_legal |> 
  rename(fips = state) |> 
  left_join(state_fips, by = "fips") |> 
  drop_na(estimate1, estimate2) |>
  mutate(
    significant_increase = factor(significant_increase, levels = c(TRUE, FALSE))
  ) 

combine_plots(plotting_df)
```