---
title: "correlation_matrix"
author: "Angelica Bailey"
date: "2025-11-19"
output: github_document
---

Identifying relationships between several variables in the dataset with correlation matrix, summarizing the strength and direction of the relationship between pair of variables.


Loading libraries
```{r}
library(tidyverse)
library(ggcorrplot)
library(patchwork)
source("./source/load_clean_brfss.R")
```


Loading in data
```{r}
brfss_data = load_clean_brfss("data/brfss_clean_2017_2024.csv")
```


Testing correlation
```{r}
#selecting variables 
correlation_vars = c("general_health_refactored", "physical_health_not_good_days","mental_health_not_good_days", "education_status", "income_level", "depressive_disorder", "life_satisfaction", "loneliness", "emotional_support")


#stratifying datasets by sex
brfss_males = brfss_data |>
  filter(sex == "Male") |>
  select(correlation_vars) |> 
  mutate_all(as.numeric)

brfss_females = brfss_data |>
  filter(sex == "Female") |>
  select(all_of(correlation_vars)) |> 
  mutate_all(as.numeric)


#creating matrices 
corr_males = 
  cor(brfss_males, 
      method = "spearman",
      use = "pairwise.complete.obs")

corr_females =
  cor(brfss_females,
      method = "spearman",
      use = "pairwise.complete.obs")

```


Visualizing matrices side by side

```{r}
#male matrix
male_plot = ggcorrplot(
  corr_males,
  type = "upper",            
  method = "square",         
  lab = TRUE,            #display coefficient values    
  lab_size = 3,              
  title = "Correlations: males") +
  theme(plot.title = element_text(hjust = 0.5)) #centering title

#females matrix
female_plot = ggcorrplot(
  corr_females, 
  type = "upper",
  method = "square",
  lab = TRUE,    
  lab_size = 3,
  title = "Correlations: females") +
  theme(plot.title = element_text(hjust = 0.5))

male_plot + female_plot

```

Next steps:
- create non-stratified correlation matrix since correlatio between female and male groups do not differ significantly 
- add age group and marital status variables

*Since education status has little correlation with other outcome variables, we can use for future analysis as predictor variable. 









