---
title: "Correlation Matrix"
author: "Angelica Bailey"
date: "2025-11-19"
output: github_document
---

Identifying relationships between several variables in the dataset with correlation matrix, summarizing the strength and direction of the relationship between pair of variables.


Loading libraries
```{r}
library(tidyverse)
library(ggcorrplot)
library(patchwork)
source(here::here("source", "load_clean_brfss.R"))
```


Loading in data
```{r}
brfss_data = load_clean_brfss(here::here("data", "brfss_clean_2017_2024.csv"))
```


Testing correlation
```{r}
#selecting variables 

correlation_vars = c("general_health","general_health_refactored", 
                     "mental_health", "mental_health_not_good_days",
                     "physical_health", "physical_health_not_good_days",
                     "poor_health", "depressive_disorder",
                     "difficulty_self_care", "michd",
                     "binge_drink", "heavy_drink", 
                     "leisure_physical_activity_last_30_days")

                          

#stratifying datasets by sex

brfss_males = brfss_data |>
  filter(sex == "Male") |>
  select(all_of(correlation_vars)) |> 
  mutate_all(as.numeric)

brfss_females = brfss_data |>
  filter(sex == "Female") |>
  select(all_of(correlation_vars)) |> 
  mutate_all(as.numeric)


#creating matrices and visualizing side by side

corr_males = 
  cor(brfss_males, 
      method = "spearman",
      use = "pairwise.complete.obs") |> 
  ggcorrplot(
    type = "upper",  
    method = "square", 
    lab = TRUE,               
    lab_size = 3, 
    title = "Correlations: males") +
  theme(plot.title = element_text(hjust = 0.5))
  
corr_females =
  cor(brfss_females,
      method = "spearman",
      use = "pairwise.complete.obs") |> 
  ggcorrplot(
    type = "upper",
    method = "square",
    lab = TRUE,
    lab_size = 3,
    title = "Correlations: females") +
  theme(plot.title = element_text(hjust = 0.5))

corr_males + corr_females
  
```



Creating non stratified matrix
```{r}
brfss_all = brfss_data |> 
  select(all_of(correlation_vars)) |> 
  mutate_all(as.numeric)

corr = 
  cor(brfss_all, 
      method = "spearman",
      use = "pairwise.complete.obs") |> 
  ggcorrplot(
    type = "upper",
    method = "square",
    lab = TRUE,
    lab_size = 3,
    title = "Correlations",
    theme(plot.title = element_text(hjust = 0.5))
  )

#to remove grid lines: theme(panel.grid.major = element_blank())

```


**Since education status has little correlation with other outcome variables, we can use it for future analysis as a predictor variable.**









