---
title: "relative_risk_eda"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
knitr::opts_chunk$set(
  fig.width = 8,
  fig.asp = 1,
  out.width = "90%"
)
library(tidyverse)
library(epitools)
source(here::here("source", "load_clean_brfss.R"))
source(here::here("source", "run_fishers_test.R"))
brfss_data = load_clean_brfss(here::here("data", "brfss_clean_2017_2024.csv.zip"))
state_fips = read_csv(here::here("data", "legal_sports_report", "state_fips.csv"))
state_fips_legalized_only = read_csv(here::here("data", "legal_sports_report", "state_legalization_dates.csv"))

outcome_vars <- c("any_physical_health_not_good_days", "any_mental_health_not_good_days",
                  "has_depressive_disorder", "has_binge_drink")
```

According to PEW research, "Young adults are more likely than older Americans to say they’ve placed a sports bet in the past year."

We are using the 4 following health outcomes:

  * *At least 1 poor mental health day*: "Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good?" 
  * *At least 1 poor physical health day*: "Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good?"
  * *Depression*: "Has a doctor or other health professional ever told you that you had a depressive disorder, including depression, major depression, dysthymia, or minor depression?"
  * *Binge Drinking:* "Considering all alcoholic beverages that you drink, on average, how many times per month do you have 5 (for men) or 4 (for women) or more drinks on one occasion?"

In 2024, what is the risk of having poor mental health outcomes based on state legalization?
Exposure: betting legalization
Outcomes: `r outcome_vars`
```{r}
phys_health_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(any_physical_health_not_good_days)) |>
  select(sb_legal, any_physical_health_not_good_days) |> 
  rename(outcome = any_physical_health_not_good_days) |> 
  run_fishers_test()

ment_health_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(any_mental_health_not_good_days)) |>
  select(sb_legal, any_mental_health_not_good_days) |> 
  rename(outcome = any_mental_health_not_good_days) |> 
  run_fishers_test()

depression_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(has_depressive_disorder)) |>
  select(sb_legal, has_depressive_disorder) |> 
  rename(outcome = has_depressive_disorder) |> 
  run_fishers_test()

binge_drink_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(has_binge_drink)) |>
  select(sb_legal, has_binge_drink) |> 
  rename(outcome = has_binge_drink) |> 
  run_fishers_test()
```
Among BRFSS participants in 2024 between the ages of 18 and 29, those who resided in states with legalized sports betting had:

  * `r round(pull(phys_health_results, risk_ratio), digits = 2)`x risk of having at least 1 day of poor physical health
  * `r round(pull(ment_health_results, risk_ratio), digits = 2)`x risk of having at least 1 day of poor mental health
  * `r round(pull(depression_results, risk_ratio), digits = 2)`x risk of having depression
  * `r round(pull(binge_drink_results, risk_ratio), digits = 2)`x risk of binge drinking 
  
compared to those residing in states where sports betting is not legal. Each of these results also show that there is a statistically significant relationship between the status of sports betting legalization and health outcomes:

  * At least 1 day of poor physical health: `r sprintf("%.2e", pull(phys_health_results, p.value))`
  * At least 1 day of poor mental health: `r sprintf("%.2e", pull(ment_health_results, p.value))`
  * Depression: `r sprintf("%.2e", pull(depression_results, p.value))`
  * Binge Drinking: `r sprintf("%.2e", pull(binge_drink_results, p.value))`

  




