---
title: "Shivalika sections"
author: "Shivalika Chavan"
date: "2025-11-28"
output: github_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(tidyverse)
library(haven)
library(lubridate)
library(here)
library(patchwork)
library(epitools)

source(here("source", "filter_brfss_data.R"))
source(here("source", "standardize_brfss_variable.R"))
source(here("source", "prop_test_functions.R"))
source(here("source", "run_fishers_test.R"))
source(here("source", "load_clean_brfss.R"))

brfss_data = load_clean_brfss(here("data", "brfss_clean_2017_2024.csv.zip"))
state_fips = read_csv(here("data", "legal_sports_report", "state_fips.csv"))

sb_rev_by_month = read_csv(here("data", "legal_sports_report", "sb_rev_by_month.csv"))
years_in_data = unique(year(pull(sb_rev_by_month, month)))

outcome_vars = c("any_physical_health_not_good_days", "any_mental_health_not_good_days", "has_depressive_disorder", "has_binge_drink")
```

# 1. Motivation
#### Provide an overview of the project goals and motivation.

# 2. Related work
Anything that inspired you, such as a paper, a web site, or something we discussed in class.

# 3. Initial questions
#### What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?

# 4. Data
#### Source, scraping method, cleaning, etc.

##### Source
Our primary data source for health outcomes is the Behavioral Risk Factor Surveillance System (BRFSS), an annual, state-based, telephone survey conducted by the CDC. We looked at all completed interviews between 2017 and 2024 (that were publicly available as of early November 2025). We chose this period as it spans the period before the Supreme Court ruling in 2018, which allowed many states to legalize sports betting. Our final dataset contained approximately `r round(nrow(brfss_data)/1000000, digits = 2)` million completed interviews across all years.

##### Cleaning
Across the years, there have been many iterations of the BRFSS survey, in terms of variable coding and questions asked. As a result, extensive cleaning and standardization was required. This process was handled with several `source` files. More detailed explanations of this process are available in `brfss_data_exploration.Rmd` and `brfss_data_cleaning.Rmd`. 

  * `standardize_brfss_variable.R`: This function renames variables that have more than one version of the same variable (e.g. `HLTHPL1` and `HLTHPL2`). Seeing that the variables differed in name only (the responses were coded in the same way), this was handled by renaming different versions to the single standardized name. (e.g. `hlthpl_standard`). 
  * `filter_brfss_data.R`: This function keeps the key variables for our study (unique survey identifiers, demographic/predictor variables, health/financial outcomes variables). It also filters for completed interviews (`dispcode = 1100`) and converts the date of the interview to a Date type variable. 
  * `load_clean_brfss_data.R`: This function is used to load the cleaned data set into R by recoding variables (e.g. `1 ~ "Male"`), and connects the date of each interview to the status of legalization in that state. This `sb_legal` treatment variable is added by merging with `state_legalization_dates.csv`.
  
Our analyses focuses on the following cleaned variable groups:

  * Treatment Variable: `sb_legal` (`1` if the interview was conducted after the state legalized sports betting (rounded to the first of the motnh), `0` if it was before, and `NA` if the state has yet to legalize sport betting) 
  * Outcome Variables: `mental_health_not_good_days`, `physical_health_not_good_days`, `depressive_disorder`, and `binge_drink`. In some cases, a binary version of these variables was used (1 if the participant has the outcome, 0 if not). 
  * Predictors: any demographic variables (sex, age group, education level, etc.)
  
# 5. Exploratory analysis
#### Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

##### Data Completeness
One of our primary concerns with data was ensuring that the key outcome variables were reported consistently from 2017 to 2024. 
```{r}
all_outcome_vars = c("general_health", "general_health_refactored", "michd",
                 "physical_health", "physical_health_not_good_days", "leisure_physical_activity_last_30_days",
                 "mental_health", "mental_health_not_good_days", "poor_health", "depressive_disorder",
                 "difficulty_self_care", "life_satisfaction", "emotional_support", "loneliness", "binge_drink",
                 "heavy_drink", "lost_reduced_employment", "financial_strain_bills", "financial_strain_utilities")

summarize_outcome_by_year = function(outcome_var, df = brfss_data) {
  
  outcome_sym = rlang::sym(outcome_var) 
  
  df |> 
    group_by(year(date), !!outcome_sym) |> 
    summarize(n = n(), .groups = "drop") |> 
    pivot_wider(
      names_from = `year(date)`, 
      values_from = n
    )
}

find_complete_years = function(outcome_var, summary_df){
  
  outcome_sym = rlang::sym(outcome_var) 
  
  summary_df |> 
    filter(!is.na(!!outcome_sym)) |> 
    summarize(
      across(
        .cols = everything(),
        .fns = ~ !all(is.na(.x)) 
        )
      )|>
    select(-!!outcome_sym) |> 
    pivot_longer(
      cols = everything(),
      names_to = "year",
      values_to = "is_complete"
      ) |> 
    summarize(
      n_complete_years = sum(is_complete)
      ) |> 
    pull(n_complete_years)
    
}

complete_outcomes = tibble(
  outcome = all_outcome_vars,
  summary = map(all_outcome_vars, summarize_outcome_by_year),
  n_complete_years = map2(all_outcome_vars, summary, find_complete_years)
  ) |> 
  arrange(n_complete_years) |> 
  filter(
    n_complete_years == 8
  ) |> 
  pull(outcome)

setdiff(all_outcome_vars, complete_outcomes)
```

Some of the outcome variables we were interested in looking at like financial strain, were not available in all years. In order to maintain a robust analysis over time, we were limited to those variables that were available over all years. Our 4 primary outcomes are available over all years. 

##### Distribution of Key Outcomes + Rationale for Binary Outcomes
```{r}
brfss_data |>
  filter(year(date) %in% c(2017, 2024)) |>
  select(year = date, mental_health, physical_health) |>
  mutate(
    year = factor(year(year)),
    mental_health = ifelse(mental_health > 30, 31, mental_health), 
    physical_health = ifelse(physical_health > 30, 31, physical_health)
  ) |>
  pivot_longer(
    cols = c(mental_health, physical_health),
    names_to = "outcome",
    values_to = "n_bad_days"
  ) |>
  ggplot(aes(x = n_bad_days, fill = year)) +
  geom_bar(position = "dodge") +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 31), labels = c(0, 7, 14, 21, 31)) +
  facet_wrap(~outcome) +
  labs(
    title = "Distribution of Poor Health Days (2017 vs. 2024)",
    x = "Number of Days of Poor Health (Past 30 Days)",
    y = "Count of Respondents"
  )
```

Given the concentration of responses at 0 days, we decided to use a binary outcome variable like `any_physical_health_not_good_days` as the outcome variable. This also follows what Couture, et. al. used in their analysis. It also simplified statistical tests that we decided to do later on. 

##### Understanding Spikes in Sports Betting Handle
In addition to the BRFSS survey, we looked at sports betting data (Total Handle) to understand the cyclic nature of the betting market that could influence mental health outcomes. The purpose is to determine if the spikes in betting activity align with major sports seasons.

###### Overall 
```{r}
sb_rev_by_month |> 
  mutate(handle_B = handle / 1e9) |> 
  ggplot(aes(x = month, y = handle_B)) + 
  geom_line() + 
  ylab("Handle (in USD, Billions)") + 
  labs(title = "Total Monthly Sports Betting Handle (USD)")
```

The plot clearly shows strong cyclical trends in total handle, suggesting that the volume of betting is highly sensitive to the sports calendar.

###### NFL Season 
```{r}
nfl_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "09", "01", sep = "-")),
    ymd(paste(years_in_data, "01", "01", sep = "-")),
    ymd(paste(years_in_data, "02", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 2 ~ "Solid", 
    TRUE ~ "Dotted" 
    ),
  
  label = case_when(
    month(date) == 9 ~ "Start of Season",
    month(date) == 1 ~ "End of Season",
    month(date) == 2 ~ "Super Bowl"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = nfl_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#D50A0A",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NFL Season", 
    values = c("Dotted" = "dotted", "Solid" = "solid"), 
    labels = c("Dotted" = "NFL Regular Season", "Solid" = "Super Bowl") 
  ) + 
  theme(legend.position="bottom")
```

The largest peak in the betting handle clearly coincides with the NFL regular season in September, and a smaller dip aligns with the season's end around the Super Bowl. 

###### NCAA March Madness
```{r}
ncaa_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "03", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 3 ~ "Solid"
    ),
  
  label = case_when(
    month(date) == 3 ~ "March Madness"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = ncaa_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#009CDE",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NCAA Events", 
    values = c("Solid" = "solid"), 
    labels = c("Solid" = "March Madness") 
  ) + 
  theme(legend.position="bottom")

```

A secondary peak in betting handle occurs during March Madness. This is important because it represents a period of intense, short-term betting interest. 

# 6. Additional analysis
#### If you undertake formal statistical analyses, describe these in detail

To test the relationship between sports betting legalization and health outcomes, we employed two statistical approaches: a proportion test to compare the prevalence of adverse outcomes before and after the legalization of sports betting, and a Fisher's Exact Test to calculate the Relative Risk (RR) of these outcomes in states where sports betting is legal vs. states where it is illegal for the most recent data in 2024. We compared outcome rates in "exposed" (sports betting is legal) and "unexposed" (sports betting is not legal) populations. 

##### Prop Tests

Our one-sided two sample proportion tests are examining whether there is a significant change (baseline is lower) in the rates of the 4 health outcomes. In the first test, for example, the null hypothesis is that the rate of a particular outcome (e.g. binge drinking) is the same in 2017 and in 2024. The alternative is that the rate of binge drinking was less in 2017 that it was in 2024. 

These tests looks at the raw change over time, regardless of when legalization occurred, providing a general benchmark. The lines show the change, and the color indicates statistical significance. 
```{r fig.width = 12, fig.asp = 0.9,  out.width = "95%"}
prop_tests_state = 
  brfss_data |> 
  filter(year(date) %in% c(2017, 2024)) |>
  select(state, date, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(test_result = map(data, run_prop_test_state)) |> 
  factor_significant()

combine_plots(prop_tests_state)
```

The second test we wanted to run used each state's specific legalization date, `sb_legal`, as the inflection point. We also repeated this test looking specifically at adults under 50 due to their higher participation in sports betting. 
```{r fig.width = 12, fig.asp = 0.9,  out.width = "95%"}
prop_tests_sb_legal = 
  brfss_data |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(test_result = map(data, run_prop_test_state_legalization)) |> 
  factor_significant()

combine_plots(prop_tests_sb_legal)
```

```{r fig.width = 12, fig.asp = 0.9,  out.width = "95%"}
prop_tests_sb_legal = 
  brfss_data |> 
  filter(age_group_5yr %in% c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49")) |> 
  select(state, sb_legal, all_of(outcome_vars)) |>
  pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "outcome_value" 
    ) |>
  group_by(state, outcome) |>
  nest() |> 
  mutate(test_result = map(data, run_prop_test_state_legalization)) |> 
  factor_significant()

combine_plots(prop_tests_sb_legal)
```

##### Relative Risk of Poor Mental Health Outcomes in Legal vs. Non-Legal states in 2024

According to PEW research, "Young adults are more likely than older Americans to say they’ve placed a sports bet in the past year."

We are using the 4 following health outcomes:

  * *At least 1 poor mental health day*: "Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good?" 
  * *At least 1 poor physical health day*: "Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good?"
  * *Depression*: "Has a doctor or other health professional ever told you that you had a depressive disorder, including depression, major depression, dysthymia, or minor depression?"
  * *Binge Drinking:* "Considering all alcoholic beverages that you drink, on average, how many times per month do you have 5 (for men) or 4 (for women) or more drinks on one occasion?"

In 2024, what is the risk of having poor mental health outcomes based on state legalization?

Exposure: betting legalization

Outcomes: `r outcome_vars`

We opted to use a Fisher's exact test in the event that one of the counts happened to be small. The goal of this test is to determine if there is a statistically significant association between status of sports betting legalization and the presence of the 4 health outcomes. 

```{r}
phys_health_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(any_physical_health_not_good_days)) |>
  select(sb_legal, any_physical_health_not_good_days) |> 
  rename(outcome = any_physical_health_not_good_days) |> 
  run_fishers_test()

phys_health_results |> pull(summary) |> knitr::kable()

ment_health_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(any_mental_health_not_good_days)) |>
  select(sb_legal, any_mental_health_not_good_days) |> 
  rename(outcome = any_mental_health_not_good_days) |> 
  run_fishers_test()

ment_health_results |> pull(summary) |> knitr::kable()

depression_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(has_depressive_disorder)) |>
  select(sb_legal, has_depressive_disorder) |> 
  rename(outcome = has_depressive_disorder) |> 
  run_fishers_test()

depression_results |> pull(summary) |> knitr::kable()

binge_drink_results = 
  brfss_data |>
  filter(year(date) == 2024, age_group_5yr %in% c("18-24", "25-29"), !is.na(has_binge_drink)) |>
  select(sb_legal, has_binge_drink) |> 
  rename(outcome = has_binge_drink) |> 
  run_fishers_test()

binge_drink_results |> pull(summary) |> knitr::kable()
```



# 7. Discussion
#### What were your findings? Are they what you expect? What insights into the data can you make?

Among BRFSS participants in 2024 between the ages of 18 and 29, those who resided in states with legalized sports betting had:

  * `r round(pull(phys_health_results, risk_ratio), digits = 2)`x risk of having at least 1 day of poor physical health
  * `r round(pull(ment_health_results, risk_ratio), digits = 2)`x risk of having at least 1 day of poor mental health
  * `r round(pull(depression_results, risk_ratio), digits = 2)`x risk of having depression
  * `r round(pull(binge_drink_results, risk_ratio), digits = 2)`x risk of binge drinking 
  
compared to those residing in states where sports betting is not legal. Each of these results also show that there is a statistically significant relationship between the status of sports betting legalization and health outcomes:

  * At least 1 day of poor physical health: `r sprintf("%.2e", pull(phys_health_results, p.value))`
  * At least 1 day of poor mental health: `r sprintf("%.2e", pull(ment_health_results, p.value))`
  * Depression: `r sprintf("%.2e", pull(depression_results, p.value))`
  * Binge Drinking: `r sprintf("%.2e", pull(binge_drink_results, p.value))`
