---
title: "Shivalika sections"
author: "Shivalika Chavan"
date: "2025-11-28"
output: github_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
library(tidyverse)
library(haven)
library(lubridate)
library(here)

# Load your custom functions
source(here("source", "filter_brfss_data.R"))
source(here("source", "standardize_brfss_variable.R"))
source(here("source", "load_clean_brfss.R"))

brfss_data = load_clean_brfss(here("data", "brfss_clean_2017_2024.csv.zip"))

sb_rev_by_month = read_csv(here("data", "legal_sports_report", "sb_rev_by_month.csv"))
years_in_data = unique(year(pull(sb_rev_by_month, month)))

```

# 1. Motivation
#### Provide an overview of the project goals and motivation.

# 2. Related work
Anything that inspired you, such as a paper, a web site, or something we discussed in class.

# 3. Initial questions
#### What questions are you trying to answer? How did these questions evolve over the course of the project? What new questions did you consider in the course of your analysis?

# 4. Data
#### Source, scraping method, cleaning, etc.

##### Source
Our primary data source for health outcomes is the Behavioral Risk Factor Surveillance System (BRFSS), an annual, state-based, telephone survey conducted by the CDC. We looked at all completed interviews between 2017 and 2024 (that were publicly available as of early November 2025). We chose this period as it spans the period before the Supreme Court ruling in 2018, which allowed many states to legalize sports betting. Our final dataset contained approximately `r round(nrow(brfss_data)/1000000, digits = 2)` million completed interviews across all years.

##### Cleaning
Across the years, there have been many iterations of the BRFSS survey, in terms of variable coding and questions asked. As a result, extensive cleaning and standardization was required. This process was handled with several `source` files. More detailed explanations of this process are available in `brfss_data_exploration.Rmd` and `brfss_data_cleaning.Rmd`. 

  * `standardize_brfss_variable.R`: This function renames variables that have more than one version of the same variable (e.g. `HLTHPL1` and `HLTHPL2`). Seeing that the variables differed in name only (the responses were coded in the same way), this was handled by renaming different versions to the single standardized name. (e.g. `hlthpl_standard`). 
  * `filter_brfss_data.R`: This function keeps the key variables for our study (unique survey identifiers, demographic/predictor variables, health/financial outcomes variables). It also filters for completed interviews (`dispcode = 1100`) and converts the date of the interview to a Date type variable. 
  * `load_clean_brfss_data.R`: This function is used to load the cleaned data set into R by recoding variables (e.g. `1 ~ "Male"`), and connects the date of each interview to the status of legalization in that state. This `sb_legal` treatment variable is added by merging with `state_legalization_dates.csv`.
  
Our analyses focuses on the following cleaned variable groups:

  * Treatment Variable: `sb_legal` (`1` if the interview was conducted after the state legalized sports betting (rounded to the first of the motnh), `0` if it was before, and `NA` if the state has yet to legalize sport betting) 
  * Outcome Variables: `mental_health_not_good_days`, `physical_health_not_good_days`, `depressive_disorder`, and `binge_drink`. In some cases, a binary version of these variables was used (1 if the participant has the outcome, 0 if not). 
  * Predictors: any demographic variables (sex, age group, education level, etc.)
  
# 5. Exploratory analysis
#### Visualizations, summaries, and exploratory statistical analyses. Justify the steps you took, and show any major changes to your ideas.

##### Data Completeness
One of our primary concerns with data was ensuring that the key outcome variables were reported consistently from 2017 to 2024. 
```{r}
outcome_vars = c("general_health", "general_health_refactored", "michd",
                 "physical_health", "physical_health_not_good_days", "leisure_physical_activity_last_30_days",
                 "mental_health", "mental_health_not_good_days", "poor_health", "depressive_disorder",
                 "difficulty_self_care", "life_satisfaction", "emotional_support", "loneliness", "binge_drink",
                 "heavy_drink", "lost_reduced_employment", "financial_strain_bills", "financial_strain_utilities")

summarize_outcome_by_year = function(outcome_var, df = brfss_data) {
  
  outcome_sym = rlang::sym(outcome_var) 
  
  df |> 
    group_by(year(date), !!outcome_sym) |> 
    summarize(n = n(), .groups = "drop") |> 
    pivot_wider(
      names_from = `year(date)`, 
      values_from = n
    )
}

find_complete_years = function(outcome_var, summary_df){
  
  outcome_sym = rlang::sym(outcome_var) 
  
  summary_df |> 
    filter(!is.na(!!outcome_sym)) |> 
    summarize(
      across(
        .cols = everything(),
        .fns = ~ !all(is.na(.x)) 
        )
      )|>
    select(-!!outcome_sym) |> 
    pivot_longer(
      cols = everything(),
      names_to = "year",
      values_to = "is_complete"
      ) |> 
    summarize(
      n_complete_years = sum(is_complete)
      ) |> 
    pull(n_complete_years)
    
}

complete_outcomes = tibble(
  outcome = outcome_vars,
  summary = map(outcome_vars, summarize_outcome_by_year),
  n_complete_years = map2(outcome_vars, summary, find_complete_years)
  ) |> 
  arrange(n_complete_years) |> 
  filter(
    n_complete_years == 8
  ) |> 
  pull(outcome)

setdiff(outcome_vars, complete_outcomes)
```

Some of the outcome variables we were interested in looking at like financial strain, were not available in all years. In order to maintain a robust analysis over time, we were limited to those variables that were available over all years. Our 4 primary outcomes are available over all years. 

##### Distribution of Key Outcomes + Rationale for Binary Outcomes
```{r}
brfss_data |>
  filter(year(date) %in% c(2017, 2024)) |>
  select(year = date, mental_health, physical_health) |>
  mutate(
    year = factor(year(year)),
    mental_health = ifelse(mental_health > 30, 31, mental_health), 
    physical_health = ifelse(physical_health > 30, 31, physical_health)
  ) |>
  pivot_longer(
    cols = c(mental_health, physical_health),
    names_to = "outcome",
    values_to = "n_bad_days"
  ) |>
  ggplot(aes(x = n_bad_days, fill = year)) +
  geom_bar(position = "dodge") +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 31), labels = c(0, 7, 14, 21, 31)) +
  facet_wrap(~outcome) +
  labs(
    title = "Distribution of Poor Health Days (2017 vs. 2024)",
    x = "Number of Days of Poor Health (Past 30 Days)",
    y = "Count of Respondents"
  )
```

Given the concentration of responses at 0 days, we decided to use a binary outcome variable like `any_physical_health_not_good_days` as the outcome variable. This also follows what Couture, et. al. used in their analysis. It also simplified statistical tests that we decided to do later on. 

##### Understanding Spikes in Sports Betting Handle
In addition to the BRFSS survey, we looked at sports betting data (Total Handle) to understand the cyclic nature of the betting market that could influence mental health outcomes. The purpose is to determine if the spikes in betting activity align with major sports seasons.

###### Overall 
```{r}
sb_rev_by_month |> 
  mutate(handle_B = handle / 1e9) |> 
  ggplot(aes(x = month, y = handle_B)) + 
  geom_line() + 
  ylab("Handle (in USD, Billions)") + 
  labs(title = "Total Monthly Sports Betting Handle (USD)")
```

The plot clearly shows strong cyclical trends in total handle, suggesting that the volume of betting is highly sensitive to the sports calendar.

###### NFL Season 
```{r}
nfl_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "09", "01", sep = "-")),
    ymd(paste(years_in_data, "01", "01", sep = "-")),
    ymd(paste(years_in_data, "02", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 2 ~ "Solid", 
    TRUE ~ "Dotted" 
    ),
  
  label = case_when(
    month(date) == 9 ~ "Start of Season",
    month(date) == 1 ~ "End of Season",
    month(date) == 2 ~ "Super Bowl"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = nfl_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#D50A0A",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NFL Season", 
    values = c("Dotted" = "dotted", "Solid" = "solid"), 
    labels = c("Dotted" = "NFL Regular Season", "Solid" = "Super Bowl") 
  ) + 
  theme(legend.position="bottom")
```

The largest peak in the betting handle clearly coincides with the NFL regular season in September, and a smaller dip aligns with the season's end around the Super Bowl. 

###### NCAA March Madness
```{r}
ncaa_event_dates = tibble(
  date = c(
    ymd(paste(years_in_data, "03", "01", sep = "-"))
    ),
  
  type = case_when(
    month(date) == 3 ~ "Solid"
    ),
  
  label = case_when(
    month(date) == 3 ~ "March Madness"
    )
  ) |> 
  filter(date >= min(sb_rev_by_month$month), date <= max(sb_rev_by_month$month))


sb_rev_by_month |>
  mutate(handle_B = handle / 1e9) |>
  ggplot(aes(x = month, y = handle_B)) +
  geom_line() +
  ylab("Handle (in USD, Billions)") +
  geom_vline(
    data = ncaa_event_dates,
    aes(xintercept = date, linetype = type),
    color = "#009CDE",
    show.legend = TRUE 
  ) +
  scale_linetype_manual(
    name = "NCAA Events", 
    values = c("Solid" = "solid"), 
    labels = c("Solid" = "March Madness") 
  ) + 
  theme(legend.position="bottom")

```

A secondary peak in betting handle occurs during March Madness. This is important because it represents a period of intense, short-term betting interest. 

# 6. Additional analysis
#### If you undertake formal statistical analyses, describe these in detail

# 7. Discussion
#### What were your findings? Are they what you expect? What insights into the data can you make?